---
title: "Piecewise Linear Segmentation by Dynamic Programming"
author: "Rainer Machn&eacute;"
date: "`r format(Sys.time(), '%d %m %Y')`"
output:
    bookdown::html_document2:
        base_format: rmarkdown::html_vignette
        toc: true
        toc_depth: 2
        fig_caption: true
bibliography: /home/raim/ref/tata.bib
vignette: >
  %\VignetteIndexEntry{Piecewise Linear Segmentation by Dynamic Programming}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, global.par=TRUE,
                      fig.path = ".", fig.pos = 'h',
                      fig.height = 2.7, fig.width = 4, fig.align = "center")
knitr::opts_knit$set(global.par = TRUE)

## redo benchmarking or use existing figure?
REDOLONG <- FALSE
times <- 100
```

```{r, include=FALSE}
par(mai=c(.5,.5,.3,.5),mgp=c(1.2,.3,0),tcl=-.25, cex.main=.75)
```

$$
\newcommand{\Ell}{\mathcal{L}}
\newcommand{\jump}{\mathcal{J}}
\newcommand{\Var}{\mathrm{Var}}
\newcommand{\Cov}{\mathrm{Cov}}
\newcommand{\lmax}{\ell_\text{max}}
\newcommand{\lmin}{\ell_\text{min}}
\newcommand{\def}{\stackrel{\mathrm{def}}{=}}
$$

TODO: FILL THIS WITH teaching/pca/cohorts14.md, teaching/pseudophase/cohorts14.md and clusterTranscriptomes_understandPCA.R

# Singular Value Decomposition and Principal Component Analysis

# The `biplot`

@Gabriel_1971


# Time Series Reconstruction: `pseudophase`.

@Alter2000,

applied for single cell analysis: @Schwabe2020, @Macosco2015


# R's `prcomp` explained

Show equivalence for prcomp, eigen and Rspectra::eigen.


# References
